---
title: |
  | Binary Logistic Regression
author: |
  | Kamarul Imran Musa
  | Assoc Prof (Epidemiology and Biostatistics)
date: '`r Sys.Date()`'
output: 
  pdf_document:
    number_sections: yes
    toc: yes
    toc_depth: 3 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

\newpage

# Introduction

We will be using examples from <http://ftp.auckland.ac.nz/software/CRAN/doc/vignettes/HSAUR/Ch_logistic_regression_glm.pdf>

We will use 5 packages

1.  the built in **stat** package - to run Generalized Linear Model. it is already loaded
2.  **HSAUR** package - to use the `plasma` dataset as an example. 
3.  **tidyverse** package - for data transformation 
4.  **broom** package - to tidy up the results 
4.  **LogisticDx** package - to do model assessment


```{r}
library(tidyverse)
library(broom)
library(HSAUR)
library(LogisticDx)
```

# Data

We will use a dataset named **plasma**.

Call data into R

```{r}
plasma <- plasma
```

Take a peek at data

```{r}
glimpse(plasma)
```

Get summary statistics

```{r}
summary(plasma)
```

The **plasma** data contain 

1.  Dependent variable = a binary variable (ESR < 20, ESR > 20)
2.  Independent variables:
    - fibrinogen - numerical variable
    - globulin -  numerical variable

Verify that ESR is a factor (categorical variable)

- $ESR>20$ is coded as 2
- $ESR<20$ is coded as 1

```{r}
str(plasma$ESR)
```

# Hands-on 1 : Estimation

Objective: To estimate the regression parameters $\hat\beta_s$

Two approaches:

1.  Univariable = 1 independent variable (covariate)
2.  Multivariable = 2 or more independent variable (covariates)

## Univariable analysis 

One dependent variable with one independent (predictor) variable.

1.  The dependent variable is ESR, a binary variable.
2.  The predictor is a numerical (continouos) variable 

We first estimate the log odds (the regression parameters, $\beta$). 

$$log\frac{Pr(ESR = high)}{1 - Pr(ESR = high)}  = \beta_0 + \beta_1(fibrinogen)$$

Estimate the parameter

```{r}
plasma_glm_1 <- glm(ESR ~ fibrinogen, data = plasma, 
                    family = binomial(link = 'logit'))
```

Get the summarized result of the model

```{r}
summary(plasma_glm_1)
```

Cleaner result in data frame format

```{r}
tidy(plasma_glm_1, conf.int = TRUE)
```

## Multivariable analysis

Unlikely only one variable (fibrinogen) relates with the outcome. Consider adding other seemingly important independent variable. We will add **globulin** to the model

Estimate the model

```{r}
plasma_glm_2 <- glm(ESR ~ fibrinogen + globulin, data = plasma,
                    family = binomial(link = 'logit'))
```

Summary of estimates

```{r}
summary(plasma_glm_2)
```

Cleaner result in data frame format

```{r}
res_mv <- tidy(plasma_glm_2, conf.int = TRUE)
res_mv
```

## Estimate the Odds Ratio

For lay person, it is difficult to interpret the log odds. It is easier to interpret - for lay person - the model using the odds ratio. 

This can be done by $\exp^{\beta_i}$

```{r}
exp(plasma_glm_1$coefficients)
```

We do not generally interpret $\hat\beta_0$. In the model, it means that if **fibrinogen** increases by 1 unit, the odds to be in ESR > 20 group, increases by 6.22 times.  

In multivariable model, the adjusted odds ratio are

```{r}
or_mv <- exp(plasma_glm_2$coefficients)
or_mv <- round(or_mv, digits = 4)
or_mv
```

Let us combine the results

```{r}
bind_cols(res_mv, adj_or = or_mv)
```

In the model, it means that if **fibrinogen** increases by 1 unit (when **globulin** is adjusted), the odds to be in ESR > 20 group, increases by 6.76 times. 

# Hands-on 2 : Inference

## Inference 

Let us examine the inferences of the model. 

This involves:

1.  checking the p-values
2.  getting the 95% confidence intervals (CI)

For the univariable model (model 1), the 95% CI for the log odds are:

```{r}
confint(plasma_glm_1)
```

Interpretation: 

With 1 unit increase in fibrinogen, the log odds for having CAD can be as small as 0.34 and as big as 4.00 at 95% confidence level. 

For the multivariable model (model 2), the $95\%$ CI for the adjusted log odds are:

```{r}
ci_mv <- confint(plasma_glm_2)
ci_mv
```

The results shows:

1.  that with 1 unit increase in fibrinogen (while adjusting for globulin), the log odds for having CAD ranged from 0.34 to 4.30 at $95\%$ CI 
2.  that with 1 unit increase in globulin (while adjusting for fibrinogen), the log odds for having CAD ranged from -0.07 to 0.43 at $95\%$ CI

Combine the results

```{r}
mv_res_beta <- bind_cols(res_mv,
                         low_ci_b = ci_mv[,1],
                         upper_ci_b = ci_mv[,2])
mv_res_beta
```


## Model comparison

Is there any difference between model 1 and model 2 at level of significance of $5\%$?

```{r}
anova(plasma_glm_2, plasma_glm_1, test = 'Chisq')
```

Not significant at $5\%$ level. But, for the practical sake, we accept the multivariable model as our model of choice

## Inference using OR

For univariable model, to estimate the odds ratio 

```{r}
exp(confint(plasma_glm_1))
```

For multivariable model, to estimate the odds ratios

```{r}
ci_or_mv <- exp(confint(plasma_glm_2))
ci_or_mv <- round(ci_or_mv, digits = 3)
ci_or_mv
```


Combine results

```{r}
or_mv
ci_or_mv
bind_cols(res_mv, adj_or = or_mv,
          l_ci_or = ci_or_mv[,1], u_ci_or = ci_or_mv[,2])
```

# Hands-on 3: Prediction

## Prediction 

We used `broom::augment()` to calculate the

1.  log odds and probability
2.  residuals
3.  hat values
4.  Cooks distance
5.  standardized residuals

## Predict the log odds

the .fitted column represents the estimated log odds for each row

```{r}
augment(plasma_glm_2)
```

## Predict the probability

the .fitted column represents the estimated probability for each row


```{r}
augment(plasma_glm_2, type.predict = "response")
```


To show the calculation for 

The first observation:

1.  fibrinogen = 2.52
2.  globulin = 38

The second observation:

1.  fibrinogen = 2.56
2.  globulin = 31

```{r}
# -12.79 + 1.91(2.52) + 0.156(38) = 
-12.7921 + 1.9104*2.52 + 0.1558*38 
-12.7921 + 1.9104*2.56 + 0.1558*31

```

Calculate the predicted probability of being in the high ESR group for all observations  

The first 6 observations have these probability (of having high ESR)

Let us confirm with manual calculation $\frac{exp^\beta}{1 + exp^\beta}$

For the first and second observations:

```{r}
exp(-2.057492) / (1 + exp(-2.057492))
exp(-3.071676) / (1 + exp(-3.071676))
```

These mean:

1.  11.3% probability of being in the high ESR group for the 1st observation 
2.  4.4% probability of being in the high ESR group for the 2nd observation

# Model fitness

```{r}
library(LogisticDx)
fit_m <- gof(plasma_glm_2)
fit_m$gof
```


# Environment

```{r}
sessionInfo()
```


